Increase of GDP and Monetary Expansion

### 2023-02-22 outputs/paper/paper.tex
final commit

---
title: Increase of GDP and Monetary Expansion
subtitle: An Analysis of Potential Factors from 16th to 19th century
author: 
  - Rae Zhang
  - Jenny Shen
  - Sarah Zhang
thanks: "Code, data and reproduction package are available at: https://github.com/JunweiZhang130/Replication-for-The-real-effects-of-monetary-expansions-evidence-from-a-large-scale-history.git and https://www.socialsciencereproduction.org/reproductions/8d4a8d6c-9f9f-44dc-bd97-9c810ed677f6/index "
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "This paper discusses the causal relationship between GDP growth and monetary expansion from the 16th to the 19th century, with the intent to identify the potential factors that contributed to economic growth during this period. Using a historical analysis approach, the paper examines multiple datasets including the nominal and real GDP of different European countries, the production of precious metals in the Americas, and the mint output of England that gave a glimpse to what potentially stimulates economic growth. The findings suggest that monetary injection played a significant role in increasing mint output and driving GDP growth during this period. Additionally, the study highlights the importance of monetary policies in promoting economic development. Overall, this paper provides insights into the factors that drove economic growth during a critical period in economic history. "
format: pdf
output: bookdown::pdf-document2
bibliography: references.bib
---

```{r}
#| echo: false
#| include: false
#| warning: false
#| message: false

library("tidyverse")
library("haven")
library("ggplot2")

```

@@

# 1 Introduction

During the early period, the discovery of massive deposits of precious metals in the Americas resulted in monetary expansions to Europe's supply. The finding of vast quantities of gold and silver in the New World had a significant influence on the world economy's subsequent history. The increase in the supply has reshaped the imports of consumer goods across Europe, and one of the results is monetary expansions. The effects of monetary policy have had major consequences, including increased economic activity, higher investment rates, and enhanced innovation. This paper explores the causal relationship of monetary policy and production of metals, and its effect on the increase of GDP, using historical data of the six European countries [@PalmaRealEffect]. Specifically, we will examine how the real GDP of European countries, Gold and Silver production, and Mint output has impacted monetary expansions. Some factors that we considered which influenced monetary expansions on a variety of outcomes, including economic activity, investment, and innovation.

We replicate the paper by Palma [@PalmaRealEffect] with a focus on the following research questions:

-   What is the trend in nominal GDP and real GDP over the period of 1500 to 1800?
-   How does gold and silver share in America contribute to the trend of monetary expansion?
-   Why is mint output increasing from the period of 1500 to 1800 and how does it contribute to the monetary expansion?

@@

The analysis found that precious metals are one of the factors that contributed to monetary expansion in the 18th century. When precious metals, such as gold and silver were used as the basis for currency, it caused an increase in demand within the country, or within international trade, which led to monetary expansion. During the period of 1500 to 1800, production of precious metal increased which contributed to many of the European country's GDP. In our reproduction of their paper, we build on their "monetary expansion" explanation by expanding on the relationship between Nominal GDP and Real GDP and the effect of precious metals in the period of 1500 to 1800.

@@

# 2 Data

All figures in the paper are replicated from the original paper; The replication material consist of two datasets: "liquidity.dta" and "datadescriptive.dta". "liquidity.dta" is the main datafile which is used for the analysis and contains all variables used in the regression analysis. "datadescriptive.dta" contains some additional variables which is used for the descriptive figures.

The datasets for this report were obtained from the data section "The Real Effects of Monetary Expansion: Evidence from a Large-scale Historical Experiment" [@PalmaRealEffect], a paper from the Review of Economic Studies. The data used by the paper was compiled by various sources that spans a very long horizon (1531--1790), The most common primary data sources for these studies are surviving account books of the landed estates formerly belonging to local government and royal administration, historical hospitals, prisons, charities, orphanages, universities, and institutions of the church, particularly monasteries and convents. GDP and prices come from the following sources: England from Broadberry et al. [@Broadberry2015], Holland from van Zanden and van Leeuwen [@vanZanden2012], Spain from ¡lvarez-Nogal and Prados de la Escosura [@AlvarezNogal2013], Italy from Malanima [@Malanima2011], Portugal from Palma and Reis (2019) and Germany from Pfister [@Pfister2020]. When the original price level was in silver it was transformed into monetary units using the information given by Karaman et al. [@Karaman2020]. Englandís nominal mint output from Challis[@Challis1992] and Palma's "Reconstruction of money supply over the long run: the case of England" and "Money and modernization in early modern England"[@Palma2018b]. With regards to the innovation component of gold arrivals, it is built using Costa et al. [@Costa2013] and Morineau[@Morineau2009] as explained in the text. The data is mainly numeric, counting the number of the amount of monetary activities in European countries.

The analysis will be carried out using the statistical programming language `R` [@citeR], using the `haven` [@citehaven] and `tidyverse` [@citetidyverse] packages. All figures in the report are generated using `ggplot2` [@citeggplot2].

```{r}
#| label: fig-bills
#| echo: false
#| warning: false
#| message: false

# Loading in datasets and library

liquidity <- read_dta(here::here("scripts/liquidity.dta"))
datadescriptive <- read_dta(here::here("scripts/datadescriptive.dta"))

```

@@

# 3 Results

## The Real and Nominal GDP over the Period of 1500 to 1800

```{r}
#| label: fig-nominalgdp
#| fig-cap: shows the nominal GDP of six countries from 1500 to 1800. *Nominal gross domestic product (GDP) is GDP given in current prices, without adjustment for inflation [@oecd_nominal_gdp_forecast]
#| echo: false
#| warning: false
#| message: false

# Replication of Figure 4
# Replication of Figure 4
color = c("#7F3C8D", 
           "#11A579", 
           "#3969AC", 
           "#F2B701", 
           "#E73F74", 
           "#80BA5A")

NominalGDP <- liquidity |>
  ggplot(aes(x = year, 
             y = Nominal_GDPindex1700100, 
             color = country)) +
  geom_line(size = 0.8) +
  scale_color_manual(values = color) +
  scale_y_continuous(limits = c(0, 300), 
                     expand = c(0, 0)) +
  labs(x = "Year", 
       y = "Nominal GDP, index(1700=100)", 
       title = "Nominal GDP for Six Countries during 1500 to 1800") +
  theme_minimal() +
  facet_wrap(~ country, ncol = 2)+
  theme(plot.title = element_text(hjust = 0.5))

NominalGDP
```

In [@fig-nominalgdp], The vertical axis shows the Nominal Gross Domestic Product (GDP) index (normalized to 1700=100), and the horizontal axis shows the year. Each graph represents one of the six countries, and the colours determine each country. The plot is intended to show the differences in the economic development of different countries over time.The impact of precious metals on GDP (Gross Domestic Product) can vary depending on several factors such as the type of metal, the production levels, and the overall demand for the metal. The data shows that there were significant differences in the level of economic development between these countries during this time period, with the United Kingdom and the Netherlands having the highest levels of nominal GDP, and Spain and Portugal having the lowest levels. \newpage

```{r}
#| label: fig-realgdp
#| echo: false
#| warning: false
#| message: false
#| fig-cap: displays the Real GDP of six European countries over the period of 1500 - 1800, *Real gross domestic product (GDP) is GDP given in constant prices and refers to the volume level of GDP [@oecd_real_gdp_forecast].

# Replication of Figure 6
color = c("#7F3C8D", 
           "#11A579", 
           "#3969AC", 
           "#F2B701", 
           "#E73F74", 
           "#80BA5A")

RealGDP <- liquidity |>
  ggplot(aes(x = year, 
             y = realGDPindex1700100, 
             color = country)) +
  geom_line(size = 0.8) +
  scale_color_manual(values = color) +
  scale_y_continuous(limits = c(0, 300), 
                     expand = c(0, 0)) +

  labs(x = "Year", 
       y = "Real GDP, index(1700=100)", 
       title = "Real GDP for Six Countries during 1500 to 1800") +
  theme_minimal() +
  facet_wrap(~ country, ncol = 2) +
  theme(plot.title = element_text(hjust = 0.5))

RealGDP

```

[@fig-realgdp] has a similar plot with [@fig-nominalgdp], the vertical axis shows the Real GDP index (normalized to 1700=100), and the horizontal axis shows the year. the response of output to a monetary expansion in the six countries from 1500 to 1800. The results indicate that the response of output varied widely between countries, with some countries experiencing a large increase in output in response to monetary expansion (e.g., the Netherlands), while others experienced little to no response (e.g., Spain). Additionally, the response of output to monetary expansion changed over time, with some countries experiencing a greater response in the earlier period of the study (e.g., Germany) and others experiencing a greater response in the later period (e.g., the United Kingdom).

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Total Goldshare through 1530 to 1790
#| label: fig-goldshare

# Replication of Figure 7
# Set color palette
mypal <- c("#1b9e77", "#d95f02")

TotalGoldShare <- datadescriptive |>
  ggplot(aes(x = year, y = goldshare, fill = "Total Gold Share")) +
  geom_bar(stat = "identity", na.rm = TRUE) +
  geom_line(aes(x = year, y = goldshare, color = "Total Gold Share"), size = 1.5) +
  labs(x = "Year", 
       y = "Total Gold Share", 
       title = "Share of Gold in Total Production through 1530 to 1790") +
  scale_x_continuous(limits = c(1530, 1790), breaks = seq(1530, 1790, by = 10)) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_fill_manual(values = mypal, guide = "none") +
  scale_color_manual(values = mypal, guide = "none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  theme(plot.title = element_text(hjust = 0.5))

TotalGoldShare
```

@@

As shown in the graph(@fig-goldshare), the mining of gold and silver culminated in the 1530s from the occurrence of Mexican and Peruvian acquisitions [@davenport_2012]. The Mexican and Peruvian conquests resulted in an outstanding increment in the production of precious metals under Spanish colonisation. However, the annual production of precious metals decreased significantly in the late sixteenth century all the way through the late seventeenth century due to factors such as shipwrecks, rivalries, and the lacking resources of gold [@woodward_bushnell]. It was not until the 1700s the share of gold and silver surges again with the new streams of Brazilian and Mexican production of precious metals [@GUERRA20041225].

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: England Mint Output from 1531 to 1790
#| label: fig-mintoutput

# Replication of Figure 12

EnglandMintOutput <- datadescriptive %>%
  ggplot(aes(x = year, y = englandmintmillions)) +
  geom_bar(stat = "identity", na.rm = TRUE) +
  labs(x = "Year", 
       y = "England Mint Output (in millions of £)", 
       title = "England Mint Output from 1531 to 1790") +
  scale_x_continuous(limits = c(1531, 1790))+
  # Adjust plot title alignment
  theme(plot.title = element_text(hjust = 0.5))
EnglandMintOutput
```

# 4 Disucussion

## 4.1 Cohort Effect in European Countries

@@

## 4.2 A Case of Monetary Expansion: England's Mint Output

[@fig-mintoutput] illustrates the English nominal mint output from 1531 - 1790 period. Mint output refers to the amount of new currency that is produced over the period of time, which can be measured using the quantity of the coins that are produced. The Mint output was a steady decrease in the issue of gold coins and sharper increase in the production of silver coins. This is primarily caused by the proportion of silver to gold coins. The mint output was relatively low in the fourteenth century, but as the picture indicates, it surged from the sixteenth century onward. During the 16th century, the world gold-silver ratio grew, as world stocks increased, the less precious metal continued to depreciate.

In the 18th century, England adopted the gold standard, which meant that the value of coins was tied to the value of gold. This led to an increase in the production of gold coins, which caused the output of the mint to fluctuate. This also explains the peaks during 1770-1780, as the mint output reached 5 million pounds. The mint output tripled in late 1700, which could be caused by several factors such as the growth of international trade. As England's economy grew and trade with other countries increased, there was a greater demand for currency to facilitate commerce [@Gould]. This supports the evidence that increase in American precious metals inevitably results in growth in European money supply, particularly for second-wave receivers like England, who obtained it mostly through trade with Spain and Portugal.

As seen in [@fig-nominalgdp] and [@fig-realgdp], the trend of mint output is linked to the trend of GDP, which can fluctuate and vary based on the economic activity of the period. While mint output of precious metals can be linked to GDP, it is worth noting that the output is not always increasing and can be volatile. This volatility is caused by fluctuations in the market, which are impacted by supply and demand within the country and in international trade. These fluctuations can lead to price increases or decreases and can be reflected in exports and imports of precious metals, which can, in turn, affect the profitability of mining companies and ultimately impact GDP.

## 4.3 How Does Mint Output and Monetary Expansion Contribute to Today's Economy?

@@

## 4.4 Bias and Ethical Concerns

The paper by Palma [@PalmaRealEffect] relies on a large dataset of economic indicators and variables, which is mostly made up of long-term historical data. The use of such data introduces several potential sources of bias that could impact the statistical analysis and interpretation of the results. One potential source of bias is selection bias. Firstly, the paper focuses on European countries and may not be generalizable to other regions. This could introduce sampling bias because the findings may not generalize to other contexts or time periods. Economic and institutional conditions in Europe during the historical period analyzed in the paper were unique and different from those in other regions and time periods. Therefore, it is possible that the results may not be applicable to other economies or time periods.

@@

# 5 Weakness

@@

# References

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Increase of GDP and Monetary Expansion},
  pdfauthor={Rae Zhang; Jenny Shen; Sarah Zhang},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Increase of GDP and Monetary Expansion\thanks{Code, data and
reproduction package are available at:
https://github.com/JunweiZhang130/Replication-for-The-real-effects-of-monetary-expansions-evidence-from-a-large-scale-history.git
and
https://www.socialsciencereproduction.org/reproductions/8d4a8d6c-9f9f-44dc-bd97-9c810ed677f6/index}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{An Analysis of Potential Factors from 16th to 19th century}
\author{Rae Zhang \and Jenny Shen \and Sarah Zhang}
\date{22 February 2023}

\begin{document}
\maketitle
\begin{abstract}
This paper discusses the causal relationship between GDP growth and
monetary expansion from the 16th to the 19th century, with the intent to
identify the potential factors that contributed to economic growth
during this period. Using a historical analysis approach, the paper
examines multiple datasets including the nominal and real GDP of
different European countries, the production of precious metals in the
Americas, and the mint output of England that gave a glimpse to what
potentially stimulates economic growth. The findings suggest that
monetary injection played a significant role in increasing mint output
and driving GDP growth during this period. Additionally, the study
highlights the importance of monetary policies in promoting economic
development. Overall, this paper provides insights into the factors that
drove economic growth during a critical period in economic history.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, enhanced, sharp corners, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

\hypertarget{introduction}{%
\section{1 Introduction}\label{introduction}}

During the early period, the discovery of massive deposits of precious
metals in the Americas resulted in monetary expansions to Europe's
supply. The finding of vast quantities of gold and silver in the New
World had a significant influence on the world economy's subsequent
history. The increase in the supply has reshaped the imports of consumer
goods across Europe, and one of the results is monetary expansions. The
effects of monetary policy have had major consequences, including
increased economic activity, higher investment rates, and enhanced
innovation. This paper explores the causal relationship of monetary
policy and production of metals, and its effect on the increase of GDP,
using historical data of the six European countries (Palma 2021).
Specifically, we will examine how the real GDP of European countries,
Gold and Silver production, and Mint output has impacted monetary
expansions. Some factors that we considered which influenced monetary
expansions on a variety of outcomes, including economic activity,
investment, and innovation.

We replicate the paper by Palma (Palma 2021) with a focus on the
following research questions:

\begin{itemize}
\tightlist
\item
  What is the trend in nominal GDP and real GDP over the period of 1500
  to 1800?
\item
  How does gold and silver share in America contribute to the trend of
  monetary expansion?
\item
  Why is mint output increasing from the period of 1500 to 1800 and how
  does it contribute to the monetary expansion?
\end{itemize}

@@

The analysis found that precious metals are one of the factors that
contributed to monetary expansion in the 18th century. When precious
metals, such as gold and silver were used as the basis for currency, it
caused an increase in demand within the country, or within international
trade, which led to monetary expansion. During the period of 1500 to
1800, production of precious metal increased which contributed to many
of the European country's GDP. In our reproduction of their paper, we
build on their ``monetary expansion'' explanation by expanding on the
relationship between Nominal GDP and Real GDP and the effect of precious
metals in the period of 1500 to 1800.

@@

\hypertarget{data}{%
\section{2 Data}\label{data}}

All figures in the paper are replicated from the original paper; The
replication material consist of two datasets: ``liquidity.dta'' and
``datadescriptive.dta''. ``liquidity.dta'' is the main datafile which is
used for the analysis and contains all variables used in the regression
analysis. ``datadescriptive.dta'' contains some additional variables
which is used for the descriptive figures.

The datasets for this report were obtained from the data section ``The
Real Effects of Monetary Expansion: Evidence from a Large-scale
Historical Experiment'' (Palma 2021), a paper from the Review of
Economic Studies. The data used by the paper was compiled by various
sources that spans a very long horizon (1531--1790), The most common
primary data sources for these studies are surviving account books of
the landed estates formerly belonging to local government and royal
administration, historical hospitals, prisons, charities, orphanages,
universities, and institutions of the church, particularly monasteries
and convents. GDP and prices come from the following sources: England
from Broadberry et al. (Broadberry et al. 2015), Holland from van Zanden
and van Leeuwen (Zanden and Leeuwen 2012), Spain from ¡lvarez-Nogal and
Prados de la Escosura (Alvarez-Nogal and Prados de la Escosura 2013),
Italy from Malanima (Malanima 2011), Portugal from Palma and Reis (2019)
and Germany from Pfister (Pfister 2020). When the original price level
was in silver it was transformed into monetary units using the
information given by Karaman et al. (Karaman, Pamuk, and
Yildirim-Karaman 2020). Englandís nominal mint output from
Challis(Challis 1992) and Palma's ``Reconstruction of money supply over
the long run: the case of England'' and ``Money and modernization in
early modern England''(Palma 2018). With regards to the innovation
component of gold arrivals, it is built using Costa et al. (Costa,
Rocha, and Sousa 2013) and Morineau(Morineau 2009) as explained in the
text. The data is mainly numeric, counting the number of the amount of
monetary activities in European countries.

The analysis will be carried out using the statistical programming
language \texttt{R} (R Core Team 2022), using the \texttt{haven}
(Wickham, Miller, and Smith 2022) and \texttt{tidyverse} (Wickham et al.
2019) packages. All figures in the report are generated using
\texttt{ggplot2} (Wickham 2016).

@@

\hypertarget{results}{%
\section{3 Results}\label{results}}

\hypertarget{the-real-and-nominal-gdp-over-the-period-of-1500-to-1800}{%
\subsection{The Real and Nominal GDP over the Period of 1500 to
1800}\label{the-real-and-nominal-gdp-over-the-period-of-1500-to-1800}}

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-nominalgdp-1.pdf}

}

\caption{\label{fig-nominalgdp}shows the nominal GDP of six countries
from 1500 to 1800. *Nominal gross domestic product (GDP) is GDP given in
current prices, without adjustment for inflation ({``Nominal {Gross}
{Domestic} {Product} {GDP}''} 2023)}

\end{figure}

In Figure~\ref{fig-nominalgdp}, The vertical axis shows the Nominal
Gross Domestic Product (GDP) index (normalized to 1700=100), and the
horizontal axis shows the year. Each graph represents one of the six
countries, and the colours determine each country. The plot is intended
to show the differences in the economic development of different
countries over time.The impact of precious metals on GDP (Gross Domestic
Product) can vary depending on several factors such as the type of
metal, the production levels, and the overall demand for the metal. The
data shows that there were significant differences in the level of
economic development between these countries during this time period,
with the United Kingdom and the Netherlands having the highest levels of
nominal GDP, and Spain and Portugal having the lowest levels. \newpage

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-realgdp-1.pdf}

}

\caption{\label{fig-realgdp}displays the Real GDP of six European
countries over the period of 1500 - 1800, *Real gross domestic product
(GDP) is GDP given in constant prices and refers to the volume level of
GDP ({``{Real GDP forecast (indicator)}''} 2023).}

\end{figure}

Figure~\ref{fig-realgdp} has a similar plot with
Figure~\ref{fig-nominalgdp}, the vertical axis shows the Real GDP index
(normalized to 1700=100), and the horizontal axis shows the year. the
response of output to a monetary expansion in the six countries from
1500 to 1800. The results indicate that the response of output varied
widely between countries, with some countries experiencing a large
increase in output in response to monetary expansion (e.g., the
Netherlands), while others experienced little to no response (e.g.,
Spain). Additionally, the response of output to monetary expansion
changed over time, with some countries experiencing a greater response
in the earlier period of the study (e.g., Germany) and others
experiencing a greater response in the later period (e.g., the United
Kingdom).

@@

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-goldshare-1.pdf}

}

\caption{\label{fig-goldshare}Total Goldshare through 1530 to 1790}

\end{figure}

@@

As shown in the graph(Figure~\ref{fig-goldshare}), the mining of gold
and silver culminated in the 1530s from the occurrence of Mexican and
Peruvian acquisitions (Davenport 2012). The Mexican and Peruvian
conquests resulted in an outstanding increment in the production of
precious metals under Spanish colonisation. However, the annual
production of precious metals decreased significantly in the late
sixteenth century all the way through the late seventeenth century due
to factors such as shipwrecks, rivalries, and the lacking resources of
gold (Woodward and Bushnell, n.d.). It was not until the 1700s the share
of gold and silver surges again with the new streams of Brazilian and
Mexican production of precious metals (Guerra 2004).

@@

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-mintoutput-1.pdf}

}

\caption{\label{fig-mintoutput}England Mint Output from 1531 to 1790}

\end{figure}

\hypertarget{disucussion}{%
\section{4 Disucussion}\label{disucussion}}

\hypertarget{cohort-effect-in-european-countries}{%
\subsection{4.1 Cohort Effect in European
Countries}\label{cohort-effect-in-european-countries}}

@@

\hypertarget{a-case-of-monetary-expansion-englands-mint-output}{%
\subsection{4.2 A Case of Monetary Expansion: England's Mint
Output}\label{a-case-of-monetary-expansion-englands-mint-output}}

Figure~\ref{fig-mintoutput} illustrates the English nominal mint output
from 1531 - 1790 period. Mint output refers to the amount of new
currency that is produced over the period of time, which can be measured
using the quantity of the coins that are produced. The Mint output was a
steady decrease in the issue of gold coins and sharper increase in the
production of silver coins. This is primarily caused by the proportion
of silver to gold coins. The mint output was relatively low in the
fourteenth century, but as the picture indicates, it surged from the
sixteenth century onward. During the 16th century, the world gold-silver
ratio grew, as world stocks increased, the less precious metal continued
to depreciate.

In the 18th century, England adopted the gold standard, which meant that
the value of coins was tied to the value of gold. This led to an
increase in the production of gold coins, which caused the output of the
mint to fluctuate. This also explains the peaks during 1770-1780, as the
mint output reached 5 million pounds. The mint output tripled in late
1700, which could be caused by several factors such as the growth of
international trade. As England's economy grew and trade with other
countries increased, there was a greater demand for currency to
facilitate commerce (Gould 1952). This supports the evidence that
increase in American precious metals inevitably results in growth in
European money supply, particularly for second-wave receivers like
England, who obtained it mostly through trade with Spain and Portugal.

As seen in Figure~\ref{fig-nominalgdp} and Figure~\ref{fig-realgdp}, the
trend of mint output is linked to the trend of GDP, which can fluctuate
and vary based on the economic activity of the period. While mint output
of precious metals can be linked to GDP, it is worth noting that the
output is not always increasing and can be volatile. This volatility is
caused by fluctuations in the market, which are impacted by supply and
demand within the country and in international trade. These fluctuations
can lead to price increases or decreases and can be reflected in exports
and imports of precious metals, which can, in turn, affect the
profitability of mining companies and ultimately impact GDP.

\hypertarget{how-does-mint-output-and-monetary-expansion-contribute-to-todays-economy}{%
\subsection{4.3 How Does Mint Output and Monetary Expansion Contribute
to Today's
Economy?}\label{how-does-mint-output-and-monetary-expansion-contribute-to-todays-economy}}

@@

\hypertarget{bias-and-ethical-concerns}{%
\subsection{4.4 Bias and Ethical
Concerns}\label{bias-and-ethical-concerns}}

The paper by Palma (Palma 2021) relies on a large dataset of economic
indicators and variables, which is mostly made up of long-term
historical data. The use of such data introduces several potential
sources of bias that could impact the statistical analysis and
interpretation of the results. One potential source of bias is selection
bias. Firstly, the paper focuses on European countries and may not be
generalizable to other regions. This could introduce sampling bias
because the findings may not generalize to other contexts or time
periods. Economic and institutional conditions in Europe during the
historical period analyzed in the paper were unique and different from
those in other regions and time periods. Therefore, it is possible that
the results may not be applicable to other economies or time periods.

@@

\hypertarget{weakness}{%
\section{5 Weakness}\label{weakness}}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-AlvarezNogal2013}{}}%
Alvarez-Nogal, Carlos, and Leandro Prados de la Escosura. 2013. {``The
Rise and Fall of Spain (1270-1850).''} \emph{The Economic History
Review} 66 (1): 1--37.

\leavevmode\vadjust pre{\hypertarget{ref-Broadberry2015}{}}%
Broadberry, Stephen, Bruce Campbell, Alexander Klein, Mark Overton, and
Bas van Leeuwen. 2015. \emph{British Economic Growth, 1270-1870}.
Cambridge University Press.

\leavevmode\vadjust pre{\hypertarget{ref-Challis1992}{}}%
Challis, C. 1992. {``Lord Hastings to the Great Silver Recoinage
1464-1699.''} In \emph{New History of the Royal Mint}, edited by C.
Challis. Cambridge University Press.

\leavevmode\vadjust pre{\hypertarget{ref-Costa2013}{}}%
Costa, L. F., M. M. Rocha, and R. Sousa. 2013. \emph{O Ouro Do Brasil}.
Imprensa Nacional-Casa da Moeda.

\leavevmode\vadjust pre{\hypertarget{ref-davenport_2012}{}}%
Davenport, Jade. 2012. {``Spanish Conquistadors and the Looting of
Mexican and Peruvian Golden Treasure.''} \emph{Mining Weekly}.
\url{https://www.miningweekly.com/article/spanish-conquistadors-and-the-looting-of-mexican-and-peruvian-golden-treasure-2012-09-07}.

\leavevmode\vadjust pre{\hypertarget{ref-Gould}{}}%
Gould, J. D. 1952. {``The Royal Mint in the Early Seventeenth
Century.''} \emph{The Economic History Review} 5 (2): 240--48.
\url{http://www.jstor.org/stable/2591058}.

\leavevmode\vadjust pre{\hypertarget{ref-GUERRA20041225}{}}%
Guerra, M. F. 2004. {``The Circulation of South American Precious Metals
in Brazil at the End of the 17th Century.''} \emph{Journal of
Archaeological Science} 31 (9): 1225--36.
https://doi.org/\url{https://doi.org/10.1016/j.jas.2004.03.018}.

\leavevmode\vadjust pre{\hypertarget{ref-Karaman2020}{}}%
Karaman, K. K., S. Pamuk, and S. Yildirim-Karaman. 2020. {``Money and
Monetary Stability in Europe, 1300--1914.''} \emph{Journal of Monetary
Economics} 115: 279--300.

\leavevmode\vadjust pre{\hypertarget{ref-Malanima2011}{}}%
Malanima, Paolo. 2011. {``The Long Decline of a Leading Economy: GDP in
Central and Northern Italy, 1300-1913.''} \emph{European Review of
Economic History} 15: 169--219.

\leavevmode\vadjust pre{\hypertarget{ref-Morineau2009}{}}%
Morineau, Michel. 2009. \emph{Incroyables Gazettes Et Fabuleux MÈtaux:
Les Retours Des TrÈsors AmÈricains díAprËs Les Gazettes Hollandaises
(XVI-XVII SiËcles)}. Cambridge University Press.

\leavevmode\vadjust pre{\hypertarget{ref-oecd_nominal_gdp_forecast}{}}%
{``Nominal {Gross} {Domestic} {Product} {GDP}.''} 2023. OECD.
\url{https://data.oecd.org/gdp/nominal-gdp-forecast.htm/\#:/~:text=Nominal/\%20gross/\%20domestic/\%20product/\%20(GDP,in/\%20the/\%20current/\%20reporting/\%20period}.
